The Iterative Signature Algorithm for Gene Expression Data

### 2009-12-08 eisa/inst/doc/EISA_tutorial.Rnw
eisa vignettes updated.

\title{The Iterative Signature Algorithm for Gene Expression data}
\author{G\'abor Cs\'ardi}
\maketitle

@@

The Iterative Signature Algorithm (ISA) is biclustering method. The
input of a biclustering method a matrix and the output is a set of
biclusters that fulfill some criteria. A bicluster is a block of the
potentially reordered input matrix.

Most commonly, this algorithm is used on microarray expression data,
to find gene sets that are coexpressed across a subset of the original
samples. In the ISA papers the biclusters are called
transription modules (TM), we will often refer them under this name in
the following.

@@

The simplest way to run ISA is to choose the two threshold parameters
and then call the \Rfunction{isa} function on the
\Rclass{ExpressionSet} object. The threshold parameters tune the size
of the modules, less stringent (i.e. smaller) values result bigger,
less correlated modules. The optimal values depend on your data and
some experimentation is needed to determine them.

Since running ISA might take a couple of minutes and the results 
depend on the random number generator used, the ISA run is commented out 
and we just load a precomputed set of modules in this vignette.
<<simple ISA>>=
thr.gene <- 2.7
thr.cond <- 1.4
set.seed(1) # to get the same results, always
# modules <- ISA(ALL, thr.gene=thr.gene, thr.cond=thr.cond)
data(ALLModulesSmall)
modules <- ALLModulesSmall
@

@@

The \Rfunction{isa} function returns an \Rclass{ISAModules} object. By
typing in its name we can get a brief summary of the results:
<<type in name of ISAModules object>>=
modules
@

@@

The \Rfunction{getNoFeatures} function returns a numeric vector, the
number of features (probesets in our case) in each module. Similarly,
\Rfunction{getNoSamples} returns a numeric vector, the number of
samples in each module. \Rfunction{pData} returns the
phenotype data of the expression set as a data frame. The
\Rfunction{getOrganism} function returns the scientific name of the
organism, \Rfunction{annotation} the name of the chip. For the former
the appropriate annotation package must be installed.
<<number of features and samples>>=
getNoFeatures(modules)
getNoSamples(modules)
colnames(pData(modules))
getOrganism(modules)
annotation(modules)
@

@@

The single bracket indexing operator can be used to restrict an
\Rclass{ISAModules} object to a subset of features and/or
samples. E.g. selecting all features that map to a gene on
chromosome~1 can be done with 
<<indexing 2>>=
chr <- get(paste(annotation(modules), sep="", "CHR"))
entrez <- sapply(mget(featureNames(modules), chr), 
                 function(x) "1" %in% x)
modules[entrez,]
@

@@

\Rfunction{getFeatureNames} lists the probesets (more precisely,
the feature names coming from the \Rclass{ExpressionSet} object) in
the modules. It returns a list, here we just print the first entry in
the list.
<<list genes in modules>>=
getFeatureNames(modules)[[1]]
@

The \Rfunction{getSampleNames} function does the same for the samples.
Again, the sample names are taken from the \Rclass{ExpressionSet}
object that was passed to \Rfunction{isa}:
<<list conditions in modules>>=
getSampleNames(modules)[[1]]
@

@@

You can also query the scores in a matrix form, that is probably
better if you need many or all of them at the same time. The
\Rfunction{getFeatureMatrix} and \Rfunction{getSampleMatrix} functions
are defined for this:
<<query all scores>>=
dim(getFeatureMatrix(modules))
dim(getSampleMatrix(modules))
@

@@

The \Rfunction{runData} function returns additional information about
the ISA run, seed the \Rfunction{isa} manual page for details.
<<run data>>=
runData(modules)
@

@@

The \Rfunction{ISAGO} function requires the annotation package on the
chip, e.g. for the ALL data, the \Rpackage{hgu95av2.db} package is
required.

@@

You can use the \Rpackage{GO.db} package to obtain more information
about the enriched GO categories.
<<GO info,cache=FALSE>>=
sigCategories(GO$BP)[[1]]
library(GO.db)
mget(sigCategories(GO$BP)[[1]], GOTERM)
@

In addition, the following functions are implemented to work on the
objects returned by \Rfunction{ISAGO}: \Rfunction{htmlReport},
\Rfunction{pvalues}, \Rfunction{geneCounts}, \Rfunction{oddsRatios},
\Rfunction{exprectedCounts}, \Rfunction{universeCounts},
\Rfunction{universeMappedCount}, \Rfunction{geneMappedCount},
\Rfunction{geneIdUniverse}. These functions do essentially the same as
they counterparts for \Rclass{GOHyperGResult} objects, see the
documentation of the \Rpackage{GOstats} package. The only difference
is, that since here we are testing a list of gene sets (=biclusters),
they calculate the results for all gene sets and usually return lists.

@@

Enrichment calculation against the KEGG pathway database goes
essentially the same way as for the Gene Ontology, this time we use
the \Rfunction{ISAKEGG} function:
<<KEGG enrichment,cache=FALSE>>=
KEGG <- ISAKEGG(modules)
KEGG
summary(KEGG, p=1e-3)[[1]]
library(KEGG.db)
mget(sigCategories(KEGG)[[1]], KEGGPATHID2NAME)
@

The functions mentioned in the Gene ontology enrichment section can be
used for KEGG, as well.

@@

The functions mentioned in the Gene ontology enrichment section can be
used for chromosome enrichment, as well.

@@

Visualizing overlapping biclusters is a challenging task. We
show simple methods that usually visualize a single bicluster 
at a time. For some of these we will use the \Rpackage{biclust} R
package.

@@

The \Rfunction{profile.plot} function has several options to set the
plot colors and styles, please see the manual for the details. This
function was inspired by the \Rfunction{parallelCoordinates} function
in the \Rpackage{biclust} package.

@@

The GO database is organized in a hierarchical fashion, in a tree-like
structure, where the broadest category sits in the root of the tree
and broader categories are subdivided into more specific
subcategories. But the GO is not exactly a tree, as the same category
can be the subcategory of more than one broader categories: e.g. the
``Golgi vesicle transport'' category is part of both
``vesicle-mediated transport'' and ``intracellular
transport''.

The \Rpackage{eisa} package provides functions to plot parts of
the GO graph that is related to a transcription module. The
\Rfunction{gograph} function creates an object that is a
representation of such a plot. Its input is a table with the GO
categories to plot and their enrichment $p$-values. (Additional
columns are silently ignored.) Here is how to use
it on the previously calculated enrichment scores:
<<goplot1>>=
goplot.2 <- gograph(summary(GO$BP, p=0.05)[[1]])
@ 
% $

\Rfunction{goplot} uses the \Rpackage{igraph} package to create a
graph with the associated meta data:
<<gograph object>>=
summary(goplot.2)
@ 
The \texttt{width} and \texttt{height} graph attributes contain the
suggested width and height of the graph in pixels, if plotted to a bitmap
device. (The graph attributes of an \Rpackage{igraph} graph can be
queried with the `\texttt{\$}' selector:
<<gograph object 2>>=
goplot.2$width
goplot.2$height
@

@@

Because the GO is not a tree, \Rfunction{gograph} ``unfolds''
it into a tree by including categories more than once, if needed. It
also abbreviates the names of the GO categories to make them fit on
the plot. The graph object contains the full name of the category as
well. The full and abbreviated names can be listed by querying the
appropriate vertex attributes of the graph. Here they are for the
first five categories:
<<goplot names>>=
V(goplot.2)$abbrv[1:5]
lapply(V(goplot.2)$desc[1:5], strwrap)
@ 
\Rfunction{gographPlot} colors the categories according to the
supplied enrichment $p$-values, the minus $\log_{10}$ $p$-value is
also added to the plot, see the bold blue numbers.

@@

It is clear that the \Sexpr{getNoFeatures(modules)[2]} genes included
in this transcription module can separate B-cell and T-cell leukemia
samples.

@@

The Rfunction{ISAHTMLModules} generates a HTML page for every single
module with the following information on it: 
\begin{itemize}
\item An expression plot of the genes and samples in the module,
  including the ISA scores. This is done by calling \Rfunction{expPlot}.
\item Overlap plot of all modules, see \Rfunction{overlapPlot}.
\item  Gene Ontology tree plots for the enriched GO terms,
  separately for the three ontologies., These are produced by
  calling \Rfunction{gograph}.
\item Tables for the enriched Gene Ontology terms, separately for
  the three ontologies.
\item A table for the enriched KEGG pathways.
\item A table for the enriched miRNA families.
\item The list of genes in the module.
\item The list of samples in the module.
\item A condition plot, see the \Rfunction{condPlot} function.
\end{itemize}
<<HTML modules>>=
ISAHTMLModules(eset=ALL, modules=modules, target.dir=htmldir,
               GO=GO, KEGG=KEGG, CHR=CHR)
@ 
The rows of the summary table generated by \Rfunction{ISAHTMLTable}
link to the module pages generated by \Rfunction{ISAHTMLModules}.

@@

The previously mentioned visualization techniques focus on a single
module only. The \Rpackage{ExpressionView} package visualizes a set of
possibly overlapping biclusters together.

@@

Using in silico data we observed that ISA has the best performance if the
input matrix is normalized (see \Rfunction{ISANormalize}). The
normalization produces two matrices: $E_r$ and
$E_c$. $E_r$ is calculated by transposing $E$ and
centering and scaling its genes (see the \Rfunction{scale} R function). $E_c$ is
calculated by centering and scaling the genes of $E$. $E_r$ is
used to calculate the sample signature of genes and $E_c$ is used
to calculate the signature of the samples.

@@

If the \Rfunction{isa} function was used to find the transcription
modules, then the robustness measure is used automatically to filter the results.
This is done by first scrambling the input matrix and then running ISA
on it. As ISA is an unsupervised algorithm it usually finds some
(although less and smaller) modules even in such a scrambled data
set. Then the robustness scores are calculated for the proper and the
scrambled modules and only (proper) modules that have a higher score
than the highest scrambled module are kept. The robustness scores are
stored in the seed data during this process, so you can check them
later:
<<robustness>>=
seedData(modules)$rob
@

@@

@@

The \Rfunction{ISA} function takes care of all steps performed during
a modular study, and for each step it uses parameters, that work reasonably
well. In some cases, however, one wants to access these steps
individually, to use custom parameters instead of the defaults.

@@

By default (i.e. if the \Rfunction{ISA} function is called) this is
performed using the \Rpackage{genefilter} package, and the default
filter is based on the inter-quantile ratio of the probesets'
expression values, a robust measure of variance. Other possible
filtering techniques include using the AffyMetrix present/absent calls
produced by the \Rfunction{mas5calls} function of the \Rpackage{affy}
package, but this requires the raw data, so in this vignette we use a
simple method based on variance and minimum expression value: only
probesets that have a variance of at least \variable{varLimit} and
that have at least \variable{kLimit} samples with expression values
over \variable{ALimit} are kept.
<<filtering>>=
varLimit <- 0.5
kLimit <- 4
ALimit <- 5
flist <- filterfun(function(x) var(x)>varLimit, kOverA(kLimit,ALimit))
ALL.filt <- ALL[genefilter(ALL, flist), ]
@

@@

In this step we match the probesets to Entrez Id and remove the ones
that don't map to any Entrez genes.
<<Entrez,cache=FALSE>>=
ann <- annotation(ALL.filt)
library(paste(ann, sep=".", "db"), character.only=TRUE)
ENTREZ <- get( paste(ann, sep="", "ENTREZID") )
EntrezIds <- mget(featureNames(ALL.filt), ENTREZ)
keep <- sapply(EntrezIds, function(x) length(x) >= 1 && !is.na(x))
ALL.filt.2 <- ALL.filt[keep,]
@

To reduce ambiguity in the interpretation of the results, we might
also want to keep only single probeset for each Entrez gene.
<<Entrez unique>>=
vari <- apply(exprs(ALL.filt.2), 1, var)
larg <- findLargest(featureNames(ALL.filt.2), vari, data=annotation(ALL.filt.2))
ALL.filt.3 <- ALL.filt.2[larg,]
@

@@

We perform the ISA iterations for our two sets of seeds separately.
<<iteration>>=
modules1 <- ISAIterate(ALL.normed, feature.seeds=random.seeds, 
                        thr.feat=1.5, thr.samp=1.8, convergence="cor")
modules2 <- ISAIterate(ALL.normed, sample.seeds=smart.seeds,
                        thr.feat=1.5, thr.samp=1.8, convergence="cor")
@

@@

The \Rfunction{ISAUnique} function eliminates the all-zero or
non-convergent input seeds and keeps only one instance of the
duplicated ones.
<<unique>>=
modules1.unique <- ISAUnique(ALL.normed, modules1)
modules2.unique <- ISAUnique(ALL.normed, modules2)
@

@@

The \Rfunction{ISAFilterRobust} function filters a set of modules by
running ISA with the same parameters on the scrambled data set and then
calculating a robustness score, both for the real modules and the ones
from the scrambled data. The highest robustness score obtained from
the scrambled data is used as a threshold to filter the real modules.
<<robust>>=
modules1.robust <- ISAFilterRobust(ALL.normed, modules1.unique)
modules2.robust <- ISAFilterRobust(ALL.normed, modules2.unique)
@

@@

\subsection{Differentially regulated modules}

@@

Let's first check, whether any of the modules can distinguish between
T-cell and B-cell ALL samples. We perform $t$-tests for both sets of
modules.
<<B vs T>>=
scores1 <- getSampleMatrix(modules1.robust)
tt1 <- colttests(scores1, as.factor(substr(type, 1, 1)))
scores2 <- getSampleMatrix(modules2.robust)
tt2 <- colttests(scores2, as.factor(substr(type, 1, 1)))
sign1 <- which(p.adjust(tt1$p.value, "holm") < 0.05)
sign2 <- which(p.adjust(tt2$p.value, "holm") < 0.05)
@

@@

As it turns out, one modules found using the smart seeds was also
found with random seeds.

@@

Let's collect the significantly enriched GO categories and KEGG
pathways.
<<dysreg enriched categories>>=
gocats <- unique(unlist(lapply(c(GO.dysreg1, GO.dysreg2), function(x) 
                               unique(unlist(sigCategories(x))))))
keggp  <- unique(unlist(lapply(list(KEGG.dysreg1, KEGG.dysreg2), function(x)
                        unique(unlist(sigCategories(x))))))
sapply(mget(gocats, GOTERM), Term)
mget(keggp, KEGGPATHID2NAME)
@

@@

From the plot it seems that these two modules are essentially the
same, one is the opposite of the others.
<<check B1 same module>>=
B1.cor <- c(cor(scores1[,which.min(tt.B1.1$p.value)],
                scores2[,which.min(tt.B1.2$p.value)]),
            cor(getFeatureMatrix(modules1.robust, 
                                 mods=which.min(tt.B1.1$p.value)),
                getFeatureMatrix(modules2.robust, 
                                 mods=which.min(tt.B1.2$p.value))))
B1.cor
@

Indeed, they are almost the same, their Pearson correlation is around
\Sexpr{round(B1.cor[1],3)} for the sample scores and
\Sexpr{round(B1.cor[2],3)} for the feature (=gene) scores.

@@

\bibliographystyle{apalike}
\bibliography{EISA}
\nocite{*}

>>>>

\title{The Iterative Signature Algorithm for Gene Expression Data}
\author{G\'abor Cs\'ardi}
\maketitle

@@

The Iterative Signature Algorithm (ISA)~\cite{sa,isa,isamod} is a
biclustering method. The input of a biclustering method is a matrix
and the output is a set of biclusters that fulfill some criteria. A
bicluster is a block of the potentially reordered input matrix.

Most commonly, biclustering algorithms are used on microarray expression data,
to find gene sets that are coexpressed across a subset of the original
samples. In the ISA papers the biclusters are called
transription modules (TM), we will often refer them under this name in
the following.

@@

The simplest way to run ISA is to choose the two threshold parameters
and then call the \Rfunction{ISA} function on the
\Rclass{ExpressionSet} object. The threshold parameters tune the size
of the modules, less stringent (i.e. smaller) values result bigger,
less correlated modules. The optimal values depend on your data and
some experimentation is needed to determine them.

Since running ISA might take a couple of minutes and the results 
depend on the random number generator used, the ISA run is commented
out from the next code block, and we just load a precomputed set of
modules that is distributed with the \Rpackage{eisa} package.
<<simple ISA,keep.source=TRUE>>=
thr.gene <- 2.7
thr.cond <- 1.4
set.seed(1) # to get the same results, always
# modules <- ISA(ALL, thr.gene=thr.gene, thr.cond=thr.cond)
data(ALLModulesSmall)
modules <- ALLModulesSmall
@

@@

The \Rfunction{ISA} function returns an \Rclass{ISAModules} object. By
typing in its name we can get a brief summary of the results:
<<type in name of ISAModules object>>=
modules
@

@@

The \Rfunction{getNoFeatures} function returns a numeric vector, the
number of features (probesets in our case) in each module. Similarly,
\Rfunction{getNoSamples} returns a numeric vector, the number of
samples in each module. \Rfunction{pData} returns the
phenotype data of the expression set as a data frame. The
\Rfunction{getOrganism} function returns the scientific name of the
organism under study, \Rfunction{annotation} the name of the chip. For
the former the appropriate annotation package must be installed.
<<number of features and samples>>=
getNoFeatures(modules)
getNoSamples(modules)
colnames(pData(modules))
getOrganism(modules)
annotation(modules)
@

@@

The single bracket indexing operator can be used to restrict an
\Rclass{ISAModules} object to a subset of features and/or
samples. E.g. selecting all features that map to a gene on
chromosome~1 can be done with 
<<indexing 2>>=
chr <- get(paste(annotation(modules), sep="", "CHR"))
chr1features <- sapply(mget(featureNames(modules), chr), 
                 function(x) "1" %in% x)
modules[chr1features,]
@

@@

\Rfunction{getFeatureNames} lists the probesets (more precisely,
the feature names coming from the \Rclass{ExpressionSet} object) in
the modules. It returns a list, here we just print the first entry.
<<list genes in modules>>=
getFeatureNames(modules)[[1]]
@

The \Rfunction{getSampleNames} function does the same for the samples.
Again, the sample names are taken from the \Rclass{ExpressionSet}
object that was passed to \Rfunction{ISA}:
<<list conditions in modules>>=
getSampleNames(modules)[[1]]
@

@@

You can also query the scores in a matrix form, that is probably
better if you need many or all of them at the same time. The
\Rfunction{getFeatureMatrix} and \Rfunction{getSampleMatrix} functions
are defined for this. The probes/samples that are not included in a
module will have a zero score by definition.
<<query all scores>>=
dim(getFeatureMatrix(modules))
dim(getSampleMatrix(modules))
@

@@

The \Rfunction{runData} function returns additional information about
the ISA run, see the \Rfunction{ISA} manual page for details.
<<run data>>=
runData(modules)
@

@@

The \Rfunction{ISAGO} function requires the annotation package of the
chip, e.g. for the ALL data, the \Rpackage{hgu95av2.db} package is
required.

@@

You can use the \Rpackage{GO.db} package to obtain more information
about the enriched GO categories.
<<GO info,cache=FALSE>>=
sigCategories(GO$BP)[[1]]
library(GO.db)
mget(sigCategories(GO$BP)[[1]][1:3], GOTERM)
@

In addition, the following functions are implemented to work on the
objects returned by \Rfunction{ISAGO}: \Rfunction{htmlReport},
\Rfunction{pvalues}, \Rfunction{geneCounts}, \Rfunction{oddsRatios},
\Rfunction{expectedCounts}, \Rfunction{universeCounts},
\Rfunction{universeMappedCount}, \Rfunction{geneMappedCount},
\Rfunction{geneIdUniverse}. These functions do essentially the same as
they counterparts for \Rclass{GOHyperGResult} objects, see the
documentation of the \Rpackage{GOstats} package. The only difference
is, that since here we are testing a list of gene sets (=biclusters),
they calculate the results for all gene sets and usually return lists.

@@

Enrichment calculation against the KEGG pathway database goes
essentially the same way as for the Gene Ontology, this time we use
the \Rfunction{ISAKEGG} function:
<<KEGG enrichment,cache=FALSE>>=
KEGG <- ISAKEGG(modules)
KEGG
summary(KEGG, p=1e-3)[[1]]
library(KEGG.db)
mget(sigCategories(KEGG, p=0.001)[[1]], KEGGPATHID2NAME)
@

The functions mentioned in the Gene ontology enrichment section
(\Rfunction{summary}, \Rfunction{pvalues}, \Rfunction{htmlReport},
etc.) can be used for chromosome enrichment, as well.

@@

The functions mentioned in the Gene ontology enrichment section
(\Rfunction{summary}, \Rfunction{pvalues}, \Rfunction{htmlReport},
etc.) can be used for chromosome enrichment, as well.

@@

Visualizing overlapping biclusters is a challenging task. We
show simple methods that usually visualize a single bicluster 
at a time. For some of these we will use the \Rpackage{biclust} R
package, \cite{biclust}.

@@

The \Rfunction{profilePlot} function has several options to set the
plot colors and styles, please see the manual for the details. This
function was inspired by the \Rfunction{parallelCoordinates} function
in the \Rpackage{biclust} package.

@@

The GO database is organized in a hierarchical fashion, in a tree-like
structure, where the broadest category sits in the root of the tree
and broader categories are subdivided into more specific
subcategories. But the GO is not a tree graph, as the same category
can be the subcategory of more than one broader categories: e.g. the
``Golgi vesicle transport'' category is part of both
``vesicle-mediated transport'' and ``intracellular
transport''.

The \Rpackage{eisa} package provides functions to plot parts of
the GO graph that are related to a transcription module. The
\Rfunction{gograph} function creates an object that is a
representation of such a plot. Its input is a table with the GO
categories to plot and their enrichment $p$-values. (Additional
columns are silently ignored.) Here is how to use
it on the previously calculated enrichment scores:
<<goplot1>>=
goplot.2 <- gograph(summary(GO$BP, p=0.05)[[1]])
@ 
% $

\Rfunction{goplot} uses the \Rpackage{igraph} package to create a
graph with the associated meta data:
<<gograph object>>=
summary(goplot.2)
@ 
The \texttt{width} and \texttt{height} graph attributes contain the
suggested width and height of the graph in pixels, if plotted to a bitmap
device. (The graph attributes of an \Rpackage{igraph} graph can be
queried with the `\texttt{\$}' selector.)
<<gograph object 2>>=
goplot.2$width
goplot.2$height
@

@@

Because the GO is not a tree, \Rfunction{gograph} ``unfolds''
it into a tree by including categories more than once, if needed. It
also abbreviates the names of the GO categories to make them fit on
the plot. The graph object contains the full names of the categories as
well. The full and abbreviated names can be listed by querying the
appropriate vertex attributes of the graph. Here they are for the
first five categories:
<<goplot names>>=
V(goplot.2)$abbrv[1:5]
lapply(V(goplot.2)$desc[1:5], strwrap)
@ 
\Rfunction{gographPlot} colors the categories according to the
supplied enrichment $p$-values, the minus $\log_{10}$ $p$-value is
also added to the plot, see the bold blue numbers.

@@

It is clear that the \Sexpr{getNoFeatures(modules)[2]} genes included
in this transcription module cannot separate B-cell and T-cell
leukemia samples. Section~\ref{sec:separators} shows good separator
modules for this data set.

@@

The \Rfunction{ISAHTMLModules} function generates a HTML page for
every single module with the following information on it: 
\begin{itemize}
\item An expression plot of the genes and samples in the module,
  including the ISA scores. This is done by calling \Rfunction{expPlot}.
\item Overlap plot of all modules, see \Rfunction{overlapPlot}.
\item  Gene Ontology tree plots for the enriched GO terms,
  separately for the three ontologies., These are produced by
  calling \Rfunction{gograph} and \Rfunction{gographPlot}.
\item Tables for the enriched Gene Ontology terms, separately for
  the three ontologies.
\item A table for the enriched KEGG pathways.
\item A table for the enriched miRNA families. (This is optional.)
\item The list of genes in the module.
\item The list of samples in the module.
\item A condition plot, see the \Rfunction{condPlot} function.
\end{itemize}
<<HTML modules>>=
ISAHTMLModules(eset=ALL, modules=modules, target.dir=htmldir,
               GO=GO, KEGG=KEGG, CHR=CHR)
@ 
The rows of the summary table generated by \Rfunction{ISAHTMLTable}
link to the module pages generated by \Rfunction{ISAHTMLModules}, if
the same \Rargument{target.dir} argument was used for both of them.

@@

The previously mentioned visualization techniques focus on a single
module at at time. The \Rpackage{ExpressionView} package visualizes a
set of possibly overlapping biclusters together.

@@

Using in silico data we observed that ISA has the best performance if the
input matrix is normalized (see \Rfunction{ISANormalize}). The
normalization produces two matrices: $E_r$ and
$E_c$. $E_r$ is calculated by transposing $E$ and
centering and scaling its expression values for each sample (see the
\Rfunction{scale} R function). $E_c$ is calculated by centering and
scaling the genes of $E$. $E_r$ is used to calculate the sample
signature of genes and $E_c$ is used to calculate the gene signature
of the samples.

@@

If the \Rfunction{ISA} function was used to find the transcription
modules, then the robustness measure is used automatically to filter the results.
This is done by first scrambling the input matrix and then running ISA
on it. As ISA is an unsupervised algorithm it usually finds some
(although less and smaller) modules even in such a scrambled data
set. Then the robustness scores are calculated for the proper and the
scrambled modules and only (proper) modules that have a higher score
than the highest scrambled module are kept. The robustness scores are
stored in the seed data during this process, so you can check them
later:
<<robustness>>=
seedData(modules)$rob
@

@@

If you analyze gene expression data, then we suggest using the
interface provided in the \Rpackage{eisa} package. For other data, use
the \Rpackage{isa2} package directly.

@@

The \Rfunction{ISA} function takes care of all steps performed in
a modular study, and for each step it uses parameters, that work reasonably
well. In some cases, however, one wants to access these steps
individually, to use custom parameters instead of the defaults.

@@

By default (i.e. if the \Rfunction{ISA} function is called) this is
performed using the \Rpackage{genefilter} package, and the default
filter is based on the inter-quantile ratio of the probesets'
expression values, a robust measure of variance.

If other filters are desired, then these can be implemented by using
the functions of the \Rpackage{genefilter} package directly.
Possible filtering techniques include using the AffyMetrix
present/absent calls produced by the \Rfunction{mas5calls} function of
the \Rpackage{affy} package, but this requires the raw data, so in
this vignette we use a simple method based on variance and minimum
expression value: only probesets that have a variance of at least
\variable{varLimit} and that have at least \variable{kLimit} samples
with expression values over \variable{ALimit} are kept.
<<filtering>>=
varLimit <- 0.5
kLimit <- 4
ALimit <- 5
flist <- filterfun(function(x) var(x)>varLimit, kOverA(kLimit,ALimit))
ALL.filt <- ALL[genefilter(ALL, flist), ]
@

@@

In this step we match the probesets to Entrez identifiers and remove
the ones that don't map to any Entrez gene.
<<Entrez,cache=FALSE>>=
ann <- annotation(ALL.filt)
library(paste(ann, sep=".", "db"), character.only=TRUE)
ENTREZ <- get( paste(ann, sep="", "ENTREZID") )
EntrezIds <- mget(featureNames(ALL.filt), ENTREZ)
keep <- sapply(EntrezIds, function(x) length(x) >= 1 && !is.na(x))
ALL.filt.2 <- ALL.filt[keep,]
@

To reduce ambiguity in the interpretation of the results, we might
also want to keep only single probeset for each Entrez gene. The
following code snipplet keeps the probeset with the highest variance.
<<Entrez unique>>=
vari <- apply(exprs(ALL.filt.2), 1, var)
larg <- findLargest(featureNames(ALL.filt.2), vari, data=annotation(ALL.filt.2))
ALL.filt.3 <- ALL.filt.2[larg,]
@

@@

We perform the ISA iterations for our two sets of seeds separately.
The two threshold parameters we use here were chosen after some
experimentation; these result modules of the ``right'' size.
<<iteration>>=
modules1 <- ISAIterate(ALL.normed, feature.seeds=random.seeds, 
                        thr.feat=1.5, thr.samp=1.8, convergence="cor")
modules2 <- ISAIterate(ALL.normed, sample.seeds=smart.seeds,
                        thr.feat=1.5, thr.samp=1.8, convergence="cor")
@

@@

The \Rfunction{ISAUnique} function eliminates the all-zero or
non-convergent input seeds and keeps only one instance of the
duplicated ones.
<<unique>>=
modules1.unique <- ISAUnique(ALL.normed, modules1)
modules2.unique <- ISAUnique(ALL.normed, modules2)
length(modules1.unique)
length(modules2.unique)
@

@@

The \Rfunction{ISAFilterRobust} function filters a set of modules by
running ISA with the same parameters on the scrambled data set and then
calculating a robustness score, both for the real modules and the ones
from the scrambled data. The highest robustness score obtained from
the scrambled data is used as a threshold to filter the real modules.
<<robust>>=
modules1.robust <- ISAFilterRobust(ALL.normed, modules1.unique)
modules2.robust <- ISAFilterRobust(ALL.normed, modules2.unique)
length(modules1.robust)
length(modules2.robust)
@

@@

\subsection{Differentially regulated modules}%
\label{sec:separators}

@@

Let's first check, whether any of the modules can distinguish between
T-cell and B-cell ALL samples. We perform $t$-tests for both sets of
modules.
<<B vs T>>=
scores1 <- getSampleMatrix(modules1.robust)
tt1 <- colttests(scores1, as.factor(substr(type, 1, 1)))
scores2 <- getSampleMatrix(modules2.robust)
tt2 <- colttests(scores2, as.factor(substr(type, 1, 1)))
sign1 <- which(p.adjust(tt1$p.value, "holm") < 0.05)
sign2 <- which(p.adjust(tt2$p.value, "holm") < 0.05)
sign1
sign2
@

@@

As it turns out, the best separator module found using the smart seeds
was also found with random seeds. This can be seen by calculating the
correlation between the separator modules found in the two experiments.
<<dysreg correlation>>=
cor(getSampleMatrix(modules1.TB), getSampleMatrix(modules2.TB))
cor(getFeatureMatrix(modules1.TB), getFeatureMatrix(modules2.TB))
@

@@

Let's collect the significantly enriched GO categories and KEGG
pathways.
<<dysreg enriched categories>>=
gocats <- unique(unlist(c(sigCategories(GO.dysreg1$BP),
                          sigCategories(GO.dysreg1$CC),
                          sigCategories(GO.dysreg1$MF),
                          sigCategories(GO.dysreg2$BP),
                          sigCategories(GO.dysreg2$CC),
                          sigCategories(GO.dysreg2$MF))))
keggp <- unique(unlist(c(sigCategories(KEGG.dysreg1), 
                         sigCategories(KEGG.dysreg2))))
sapply(mget(gocats, GOTERM), Term)
mget(keggp, KEGGPATHID2NAME)
@

@@

From the plot it seems that these two modules are essentially the
same.
<<check B1 same module>>=
B1.cor <- c(cor(scores1[,which.min(tt.B1.1$p.value)],
                scores2[,which.min(tt.B1.2$p.value)]),
            cor(getFeatureMatrix(modules1.robust, 
                                 mods=which.min(tt.B1.1$p.value)),
                getFeatureMatrix(modules2.robust, 
                                 mods=which.min(tt.B1.2$p.value))))
B1.cor
@

Indeed, they are almost the same, their Pearson correlation is
\Sexpr{round(B1.cor[1],3)} for the sample scores and
\Sexpr{round(B1.cor[2],3)} for the feature (=gene) scores.

@@

\bibliographystyle{apalike}
\bibliography{EISA}
